Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-03-05 paper/paper.tex
on section 4 (up to 4.3)

\tname{} is a technique to synthesize rules for rule-based
\nids~(\eg{}, Snort~\cite{snort} and Suricata~\cite{suricata}). The
goal is to automatically create rules that capture \emph{only} the
malicious traffic.

\vspace{1ex}
\noindent\textbf{Overview.}~\tname\ takes on input one or more
examples of attacks and produces on output a ranked list of
\emph{plausible rules}---\ie{}, rules that isolate the malicious
traffic produced by the attacks from the benign
traffic. Figure~\ref{fig:overview} shows the workflow of \tname{} as a
pipeline of three components, appearing numbered in the figure.
Public data sets of benign
traffic~\cite{tcpreplay,stratosphere-normal} and rules
exist~\cite{emerging-threats-open}. Consequently, a user of \tname{}
only needs to provide examples of attacks on input.

@@

The seed rule produced in the first step satisfies the criterion to
only capture positive traffic. However, that rule is overconstrained.
It could include options that are irrelevant to capturing the attack
and could lead to false negatives---the \nids\ may be unable to
capture manifestations of the same attack whose messages do not
exhibit the data satisfying irrelevant options.

@@

\tname{} leverages public data sets of benign traffic~\cite{tcpreplay}
and additional examples of the attack to guide the search for more
general rules to address the overffiting issue. To that end, \tname{}
\emph{minimizes} the rules produced in the previous step, preserving
the invariant of \emph{only} capturing the positive malicious
traffic. We say that a rule is minimal if discarding any option
results in the capture of negative (\ie{}, benign) traffic.  The
procedure we propose in this step builds on the observation that all
options from the rule produced in the previous step are satisifed by
the traffic produced by the attack. Consequently, any subset of these
options also captures the attack. The rule creation procedure
iteratively discards options from visited rules---derived from the
seed rule---using the negative traffic to guide the search for minimal
rules.

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algnewcommand\And{\textbf{and} }%
\setlength{\textfloatsep}{3pt}% Remove \textfloatsep
\begin{algorithm}[t!]
  %\noindent\hspace{-2ex}
  \flushleft{}\textbf{INPUT:} (overconstrained) seed rule $sr$, bening traffic
  $\mathit{bTr}$, malli-\\
  cious traffic $\mathit{mTr}$\\
  %\noindent\hspace{-27.5ex} 
  \textbf{OUTPUT:} Set of minimized rules $\mathcal O$\\
%\vspace{-3ex}
\captionof{algorithm}{Rule creation algorithm}\label{algo1}
\begin{algorithmic}[1]
  \State $\mathcal R \gets \{\mathit{sr}\}$\Comment{initial population of candidate solutions.}\label{line-init}
  \State $\mathcal O \gets \emptyset$
  \Do\label{start-do-while}
      \State $\mathcal R' \gets \emptyset$ \Comment{new iteration. reset.}\label{init-rprime}
      \ForEach {$ro \in \mathcal R $}\Comment{analyze each candidate rule.}\label{start-foreach}
           \For {$i \gets 1$ to $N$}\Comment{mutate each rule $N$ times.}\label{start-for}
               \State $r \gets copy(ro)$\label{copy}
               \State $j\gets \mathit{rand}(0,\mathit{len}(r))$\label{choose-index}
               \State $r[j]\gets 0$\Comment{discard option $j$ in $r$}\label{delete-option}
               \If{$\mathit{isNew(r)} \And\mathit{isPlausible}(r, \mathit{bTr})$} 
                   \State $\mathcal R' \gets \mathcal R' \cup \{r\}$\label{line-progress}
               \EndIf
           \EndFor\label{end-for}
       \EndFor\label{end-foreach}
       \State $\mathcal R \gets \mathcal R'$\Comment{update new generation}\label{define-next-gen}
       \State $\mathcal O \gets \mathcal O \cup \mathcal  R$\Comment{remember all rules}\label{output-accum}
  \doWhile{$\mathcal R\not= \{\}$}\Comment{repeat until reaching a
    fixpoint}\label{end-do-while}\\
  \Return $\mathit{filter(\mathcal O, mTr)}$\label{filtering}
\end{algorithmic}
\end{algorithm}

Algorithm~\ref{algo1} shows the pseudocode for deriving rules from the
seed rule produced at the reverse engineering step. The algorithm
takes as input an overconstrained rule $sr$, produced by the reverse
engineering step, and produces on output a set of alternative rules to
$sr$ that captures every instance of the attack ($\mathit{mTr}$), but
does not capture any message from the benign traffic ($\mathit{bTr}$).

Line~\ref{line-init} initializes the current population of individuals
(\ie{}, rules) $\mathcal{R}$ with the seed rule $sr$. We encode a rule
as a bitvector where each index in the vector represents one option;
the value assigned at a given position indicates whether or not the
option is present in the rule. One iteration of the outer loop
(lines~\ref{start-do-while}-\ref{end-do-while}) processes the current
population of individuals $\mathcal R$ and creates the next generation
$\mathcal R'$. Line~\ref{init-rprime} initializes the population of
individuals for the next generation. The inner loop
(lines~\ref{start-foreach}--\ref{end-foreach}) iterates through each
candidate rule in $\mathcal R$. The algorithm mutates each of these
rules for $N$ times, discarding one option of the rule each
time. Line~\ref{copy} assigns a copy of the bitvector, encoding
$\mathcal R$, to variable $r$. Function $\mathit{len}$ returns the
length of a vector and function $\mathit{rand}$ randomly chooses one
integer value in the range defined by its parameters. So, the
expression $\mathit{rand(0, len(r))}$, appearing at
line~\ref{choose-index}, returns an index of the option in $r$ to
delete at line~\ref{delete-option}. The call to $\mathit{isNew}$
checks if $r$ has not been previously generated through another path
whereas the call to $\mathit{isPlausible}$ checks if the rule only
captures positive traffic, \ie{} no message present in $\mathit{bTr}$
is captured by $r$. If $r$ satisfies both criteria, it is added to
$\mathcal R'$, which denotes the next generation of individuals in the
search. After processing every candidate rule in $\mathcal R$, the
next generation of individuals is defined (line~\ref{define-next-gen})
and the set of discovered rules is updated (at
line~\ref{output-accum}).

The algorithm terminates when no rule in the current generation of
individuals can be further reduced without violating the property that
a candidate solution should only capture positive traffic. This means
that execution of one iteration has not reached the statement declared
at line~\ref{line-progress}, which adds one element to the set
$\mathcal{R'}$. As $\mathcal{R'}$ is initialized empty that set will
remain empty at the end of an iteration of the outer loop.

Note that the algorithm produces no rules in the last iteration. Rules
with a single option are produced in the second-to-last iteration. In
other iterations, rules with more than one option are
produced. Conceptually, the search induces a search tree with the seed
rule $sr$ as the root, single-option rules as child nodes, and rules
with more options as internal nodes. The edges of the tree denote the
ancestor relation between two rules.

It is also worth noting that the algorithm reports every rule in the
tree instead of only those that are surely minimal, \ie{} rules with
one option. The reason is that the the traffic used to check
plausibility is limited by definition---reporting only minimal rules
with respect to that traffic could discard the correct rule. In
summary, this algorithm reports many rules that are plausible by using
the benign traffic to reduce the overconstrained rule produced at the
\reveng\ step. In the Raking step (Section~\ref{sec:ranking}),
\tname{} ranks the rules this algorithm reports by their similarity to
existing trusted rules. Intuitively, rules that do not look like those
trusted rules rank low.

\begin{proposition}
  \vspace{-1ex}  
  The rule generation algorithm reports rules that capture only the
  positive traffic.
\end{proposition}
\begin{proof}
  \vspace{-1.5ex} Every rule created during the search captures the
  attack, provided as example on step one, as every rule contains a
  proper subset of the options from the original rule and all those
  options are satisfied by the traffic produced by the attack. By
  definition of function $\mathit{isPlausible}$, these rules also
  capture \emph{only} the positive traffic. If negative traffic is
  captured, a rule is not added to $\mathcal R'$. Consequently, it
  cannot be an element of $\mathcal O$ and reported on the
  output. Finally, only the rules in $\mathcal O$ that capture all the
  instances of the attack (\ie{}, variants) are reported on output.
  \vspace{-1ex}    
\end{proof}

>>>>

\tname{} is a technique to synthesize rules for rule-based
\nids~(\eg{}, Snort~\cite{snort} and Suricata~\cite{suricata}). The
goal is to automatically create rules that capture an attack of
interest.

\vspace{1ex}
\noindent\textbf{Overview.}~\tname\ takes on input one or more
examples of the same attack and produces on output a ranked list of
\emph{plausible rules}---\ie{}, rules that isolate the malicious
traffic produced by the attacks from the benign
traffic. Figure~\ref{fig:overview} shows the workflow of \tname{} as a
pipeline of three components, appearing numbered in the figure.
Public data sets of benign
traffic~\cite{tcpreplay,stratosphere-normal} and rules
exist~\cite{emerging-threats-open}. Consequently, a user of \tname{}
only needs to provide examples of attacks on input.

@@

The seed rule produced in the first step satisfies the criterion to
capture the traffic produced by the example attacks. However, that
rule is overconstrained.  It may include options that are irrelevant
to capture the attack and lead to false negatives---the \nids\ may be
unable to capture manifestations of the same attack whose messages do
not exhibit the data satisfying irrelevant options. For example, the
seed rule produced after the Reverse Engineering step on the Ping Scan
example from Section~\ref{sec:active-recon} contains options
\CodeIn{icode:0}, \CodeIn{icmp\_id:23570}, and
\CodeIn{icmp\_seq:3439}, which are irrelevant to capture that attack.

@@

To address the overffiting issue, \tname{} leverages public data sets
of benign traffic~\cite{tcpreplay} and additional examples of the
attack to guide the search for more general rules. To that end,
\tname{} \emph{minimizes} the rules produced in the previous step,
preserving the invariant of \emph{only} capturing the positive
malicious traffic. We say that a rule is minimal if discarding any
option results in the capture of negative (\ie{}, benign) traffic.
The procedure we propose in this step builds on the observation that
all options from the rule produced in the previous step are satisifed
by the traffic produced by the attack. Consequently, any subset of
these options also captures the attack. The rule creation procedure
iteratively discards options from visited rules---derived from the
seed rule---using the negative traffic to guide the search for minimal
rules.

\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}%
\algnewcommand\algorithmicforeach{\textbf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}
\algnewcommand\And{\textbf{and} }%
\setlength{\textfloatsep}{3pt}% Remove \textfloatsep
\begin{algorithm}[t!]
  %\noindent\hspace{-2ex}
  \flushleft{}\textbf{INPUT:} (overconstrained) seed rule $sr$, pcap
  files for bening traffic\\
  ($\mathit{bTr}$), pcap files for mallicious traffic ($\mathit{mTr}$)\\
  %\noindent\hspace{-27.5ex} 
  \textbf{OUTPUT:} Set of minimized rules $\mathcal O$\\
%\vspace{-3ex}
\captionof{algorithm}{Rule creation algorithm}\label{algo1}
\begin{algorithmic}[1]
  \State $\mathcal R \gets \{\mathit{sr}\}$\Comment{initial population of candidate solutions.}\label{line-init}
  \State $\mathcal O \gets \emptyset$
  \Do\label{start-do-while}
      \State $\mathcal R' \gets \emptyset$ \Comment{new iteration. reset.}\label{init-rprime}
      \ForEach {$ro \in \mathcal R $}\Comment{analyze each candidate rule.}\label{start-foreach}
           \For {$i \gets 1$ to $N$}\Comment{mutate each rule $N$ times.}\label{start-for}
               \State $r \gets copy(ro)$\label{copy}
               \State $j\gets \mathit{rand}(0,\mathit{len}(r))$\label{choose-index}
               \State $r[j]\gets 0$\Comment{discard option $j$ in $r$}\label{delete-option}
               \If{$\mathit{isNew(r)} \And\mathit{isPlausible}(r, \mathit{bTr})$} 
                   \State $\mathcal R' \gets \mathcal R' \cup \{r\}$\label{line-progress}
               \EndIf
           \EndFor\label{end-for}
       \EndFor\label{end-foreach}
       \State $\mathcal R \gets \mathcal R'$\Comment{update new generation}\label{define-next-gen}
       \State $\mathcal O \gets \mathcal O \cup \mathcal  R$\Comment{remember all rules}\label{output-accum}
  \doWhile{$\mathcal R\not= \{\}$}\Comment{repeat until reaching a
    fixpoint}\label{end-do-while}\\
  \Return $\mathit{filter(\mathcal O, mTr)}$\label{filtering}
\end{algorithmic}
\end{algorithm}

Algorithm~\ref{algo1} shows the pseudocode for deriving rules from the
seed rule produced at the Reverse Engineering step. The algorithm
takes as input an overconstrained rule $sr$ and pcap files~\cite{pcap}
to reconstruct the benign ($\mathit{bTr}$) and malicious traffic
($\mathit{mTr}$). The algorithm produces on output a set of
alternative rules to $sr$ that captures every instance of the attack
(as per $\mathit{mTr}$), but does not capture any message from the
benign traffic (as per $\mathit{bTr}$).

Line~\ref{line-init} initializes the current population of individuals
(\ie{}, rules) $\mathcal{R}$ with the seed rule $sr$. We encode a rule
as a bitvector where each index in the vector represents one option;
the value assigned at a given position indicates whether or not the
option is present in the rule. One iteration of the outer loop
(lines~\ref{start-do-while}-\ref{end-do-while}) processes the current
population of individuals $\mathcal R$ and creates the next generation
$\mathcal R'$. Line~\ref{init-rprime} initializes the population of
individuals for the next generation. The inner loop
(lines~\ref{start-foreach}--\ref{end-foreach}) iterates through each
candidate rule in $\mathcal R$. The algorithm mutates each of these
rules for $N$ times, discarding one option of the rule each
time. Line~\ref{copy} assigns a copy of the bitvector, encoding
$\mathit{ro}$, to variable $r$. Function $\mathit{len}$ returns the
length of a vector and function $\mathit{rand}$ randomly chooses one
integer value in the range defined by its parameters. So, the
expression $\mathit{rand(0, len(r))}$, appearing at
line~\ref{choose-index}, returns an index of the option in $r$ to
delete at line~\ref{delete-option}. The call to $\mathit{isNew}$
checks if $r$ has not been previously generated in another iteration
whereas the call to $\mathit{isPlausible}$ checks if the rule only
captures positive traffic, \ie{}, it checks that no message present in
$\mathit{bTr}$ is captured by $r$. If $r$ satisfies both criteria, it
is added to $\mathcal R'$, which denotes the next generation of
individuals in the search. After processing every candidate rule in
$\mathcal R$, the next generation of individuals is defined
(line~\ref{define-next-gen}) and the set of discovered rules is
updated (at line~\ref{output-accum}).

The do-while loop terminates when no rule in the current generation of
individuals can be further reduced without violating the invariant
that a candidate solution should only capture positive traffic. This
means that execution of one iteration of the do-while loop has not
reached the statement declared at line~\ref{line-progress}, which adds
one element to the set $\mathcal{R'}$. As $\mathcal{R'}$ is
initialized empty, that set will remain empty at the end of the
execution of the loop from
lines~\ref{start-foreach}-\ref{end-foreach}. The algorithm returns the
rules from $\mathcal O$ that capture \emph{all} example attacks
provided on input (line~\ref{filtering}).

Note that the algorithm produces no rules in the last iteration. Rules
with a single option are produced in the second-to-last iteration. In
other iterations, rules with more than one option are
produced. Conceptually, the search induces a search tree with the seed
rule $sr$ as root, single-option rules as child nodes, and rules with
more options as internal nodes. The edges of the tree denote the
ancestor relation between a pair of rules. It is also worth noting
that the algorithm reports every rule in the tree instead of only
those with one option. The reason is that the traffic used to check
plausibility is limited by definition---reporting only minimal rules
with respect to that traffic could discard the correct rule.

In summary, this algorithm reports many rules that are plausible by
using the benign traffic to reduce the overconstrained rule produced
at the \reveng\ step. In the Raking step (Section~\ref{sec:ranking}),
\tname{} ranks the rules this algorithm reports by their similarity to
existing trusted rules. Intuitively, rules that do not look like those
trusted rules rank low.

\begin{proposition}
  \vspace{-1ex}  
  The rule generation algorithm reports rules that capture only the
  positive traffic.
\end{proposition}
\begin{proof}
  \vspace{-1.5ex} Every rule created during the search captures the
  attack provided as example on step one as every created rule
  contains a proper subset of the options from the original rule and
  all those options are, by definition, satisfied by the traffic
  produced by the original attack. In addition, from the definition of
  function $\mathit{isPlausible}$, reported rules capture \emph{only}
  the positive traffic. If negative traffic is captured, a rule is not
  added to $\mathcal R'$. Consequently, it cannot be an element of
  $\mathcal O$ and reported on the output. Finally, from the
  definition of function $\mathit{filter}$, only the rules in
  $\mathcal O$ that capture all the instances of the attack (\ie{},
  variants) are reported on output.
  \vspace{-1ex}    
\end{proof}
